---
title: "Running MOFAcell in lupus"
author: "Ricardo Ramirez"
date: "2023-05-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Additional validation data set

As a simple example, we will look at ~25k PBMCs from 8 pooled patient lupus samples, each before and after IFN-beta stimulation (Kang et al., 2018; GSE96583). Note that by focusing on PBMCs, for the purpose of this tutorial, we assume that coordinated events occur among them.

# Importing MOFAcellular

```{r setup, include=FALSE, message=FALSE}
library(MOFAcellulaR)
library(tidyverse)
```

```{r}
reticulate::use_condaenv(condaenv = "/Users/ricardoramirez/opt/miniconda3/envs/MOFA")
```

## Reading preprocessed data

```{r, message=FALSE}
pbcounts <- read_csv("./data_lupus/pb_data.csv") %>%
  dplyr::rename("pb_sample" = "...1") %>%
  dplyr::mutate(pb_sample = gsub("&","_", pb_sample)) %>%
  tibble::column_to_rownames("pb_sample") %>%
  as.matrix() %>%
  t()
```

```{r}
coldata <- read_csv("./data_lupus/cell_counts.csv") %>%
  dplyr::rename("index" = "...1",
                "donor_id" = "sample",
                "cell_counts" = "counts",
                "cell_type" = "cell_abbr") %>% 
  dplyr::select(-index) %>%
  dplyr::mutate(donor_id = gsub("&","_", donor_id)) %>%
  dplyr::mutate(col_name = paste0(donor_id, "_", cell_type)) %>%
  tibble::column_to_rownames("col_name")

coldata <- coldata[colnames(pbcounts),]
```

## What's in the data

16 samples, 8 cell-types

```{r}
coldata$donor_id %>% unique() %>% length()
coldata$cell_type %>% unique() %>% length()
```

## Pick cell types with more than 25 cells

```{r}
cts <- coldata %>%
  dplyr::filter(cell_counts >= 25) %>%
  dplyr::group_by(cell_type) %>%
  summarize(nsamples = n()) %>%
  dplyr::filter(nsamples > 10) %>%
  pull(cell_type)
```

## Running MOFAcell preprocessing

```{r}
multiview_dat <- MOFAcellulaR::create_init_exp(counts = pbcounts,  
                                               coldata = coldata) %>%
  MOFAcellulaR::filt_profiles(pb_dat = .,
                          cts = cts,
                          ncells = 25,
                          counts_col = "cell_counts", # This refers to the column name in testcoldata where the number of cells per profile was stored
                          ct_col = "cell_type") %>%
  MOFAcellulaR::filt_gex_byexpr(pb_dat_list = .,
                                min.count = 10,
                                min.prop = 0.25) %>%
  MOFAcellulaR::tmm_trns(pb_dat_list = .,
                         scale_factor = 1000000) %>%
  MOFAcellulaR::filt_gex_byhvg(pb_dat_list = .,
                               prior_hvg = NULL,
                               var.threshold = 0) %>%
  MOFAcellulaR::pb_dat2MOFA(pb_dat_list = .)
```

```{r}
multiview_dat %>% 
  dplyr::select(view, feature) %>% 
  unique() %>% 
  group_by(view) %>% 
  summarize(n_genes = n()) %>% 
  pull(n_genes) %>%
  median()
```

## Running MOFA

```{r}
MOFAobject <- MOFA2::create_mofa(multiview_dat)

data_opts <- MOFA2::get_default_data_options(MOFAobject)
train_opts <- MOFA2::get_default_training_options(MOFAobject)
model_opts <- MOFA2::get_default_model_options(MOFAobject)

# This avoids the regularization of multicellular programs per cell type.
# This avoids less sparse gene weights
model_opts$spikeslab_weights <- FALSE 

# Define the number of factors needed
model_opts$num_factors <- 4

# Prepare MOFA model:
MOFAobject <- MOFA2::prepare_mofa(object = MOFAobject,
                           data_options = data_opts,
                           model_options = model_opts,
                           training_options = train_opts)
  
outfile <- file.path("./results/lupus/lupusmodel.hdf5")

model <- MOFA2::run_mofa(MOFAobject, outfile)
```

## Exploring model

```{r}
metadata <- read_csv("./data_lupus/meta_data.csv") %>%
  dplyr::rename("index" = "...1") %>%
  dplyr::select(-index) %>%
  dplyr::mutate(sample = gsub("&","_", sample)) 

head(metadata)
```

## 2D plot

```{r, fig.height=2, fig.width=2.5}
UMAP_embedding <- MOFAcellulaR::plot_sample_2D(model = model,
                                               method = "UMAP",
                                               metadata = metadata,
                                               sample_id_column = "sample",
                                               color_by = "condition")
```

```{r}
stim_assoc <- MOFAcellulaR::get_associations(model = model,
                                      metadata = metadata,
                                       sample_id_column = "sample",
                                       test_variable = "condition",
                                       test_type = "categorical",
                                      categorical_type = "non-parametric",
                                       group = FALSE)

stim_assoc
```

```{r}
assoc_list <- list("stimuli" = stim_assoc)

col_list <- list(condition = c("stim" = "#FF6666", 
                                   "ctrl" = "#4169E1"))

cplx_hmap <- plot_MOFA_hmap(model = model,
                group = FALSE,
                metadata = metadata,
                sample_id_column = "sample",
                sample_anns = c("condition"),
                assoc_list = assoc_list,
                col_rows = col_list)

draw(cplx_hmap)

pdf("./results/lupus/lupus_summary.pdf", height = 4.7, width = 3.4)

draw(cplx_hmap)

dev.off()
```

## paper results

Total explained variance

```{r}
model@cache$variance_explained$r2_total$single_group %>% mean()
```

```{r}
model@cache$variance_explained$r2_total$single_group %>% sort()
```
First factor

```{r}
model@cache$variance_explained$r2_per_factor$single_group[1, ] %>% mean()
```







