---
title: "Estimation of multicellular factors using MOFA"
author:
- name: "Ricardo O. Ramirez Flores"
  affiliation: "Heidelberg University, Faculty of Medicine"
  email: "roramirezf@uni-heidelberg.de"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r, message=FALSE, warning=FALSE}
library(MOFA2)
library(HDF5Array)
library(ComplexHeatmap)
library(circlize)
library(MOFAcellulaR)
library(tidyverse)
library(ggpubr)
#source("./MOFAcell/MOFAcellprep.R")
#source("./MOFAcell/factorutils.R")
```

```{r, echo=FALSE}
reticulate::use_condaenv(condaenv = "/Users/ricardoramirez/opt/miniconda3/envs/MOFA")
```

# Introduction

MOFA can be used to identify coordinated gene expression programs of multiple cell types from single cell RNA-seq data. Similarly as other computational tools such as scITD or DIALOGUE, MOFA factors can be interpreted as shared gene programs between cell types, whose variability is associated to different covariates in complex experimental designs.

In this markdown we will estimate "multicellular factors" from a single-cell RNA-seq dataset of human myocardial infarction patients and healthy donors from a study by Kuppe, Ramirez Flores, Li, et al.

# Pseudobulk expression profiles and meta-data

In this section we import a SingleCellExperiment containing the profiles of 27 combinations of patient and physiological samples at different time points after myocardial infarction as well as control samples.

These specimens were classified in three major groups based on histological observations, nevertheless these are only approximations of a class

```{r}
# Defining the patient annotations -----------------------------------------------------------------
annotation_names <- tibble(patient_group = c("group_1", "group_2", "group_3"),
                           patient_group_name = c("myogenic-enriched", "ischemic-enriched", "fibrotic-enriched"))

batch_info <- read_csv("./data_MI/snrna_batch_ann.csv") %>%
  select(orig.ident, batch) %>%
  unique()

sample_dict <- read_csv("./data_MI/rna_patient_anns_revisions.csv",show_col_types = F) %>%
  left_join(annotation_names, by = "patient_group") %>%
  dplyr::select(-patient_group) %>%
  dplyr::rename("patient_group" = patient_group_name) %>%
  dplyr::mutate(patient_group = factor(patient_group, 
                                       levels = c("myogenic-enriched", "ischemic-enriched", "fibrotic-enriched"))) %>%
  #dplyr::select(-sample_id) %>%
  unique()

sample_dict_red <- sample_dict %>%
  left_join(batch_info, by = c("sample_id" = "orig.ident")) %>%
  dplyr::select(patient_region_id, major_labl, patient_group, batch) %>%
  unique() 
```

```{r, eval = F, echo= TRUE}
mi_pb <- loadHDF5SummarizedExperiment("./data_MI/integrated_rnasamples_ann_sce/")

ids <- colData(mi_pb) %>% 
  as.data.frame() %>% 
  left_join(sample_dict, by = c("orig.ident" = "sample_id")) %>%
  dplyr::select(cell_type, patient_region_id) %>%
  DataFrame()

mi_pb <- scuttle::sumCountsAcrossCells(x = mi_pb,
                                       ids  = ids)

assay(mi_pb, "counts") <- assay(mi_pb, "sum")
```

```{r, eval=F, echo=F}
saveRDS(mi_pb, "./data_MI/mi_pb.rds")
```

```{r, eval=T, echo=F}
mi_pb <- readRDS("./data_MI/mi_pb.rds")
```

Finally prepare the data for the universal input of MOFAcell preprocessing

```{r, eval=T, echo=F}
counts <- assay(mi_pb, "counts")
coldata <- colData(mi_pb) %>%
  as.data.frame() %>%
  dplyr::rename("donor_id" = "patient_region_id",
         "cell_counts" = "ncells") %>%
  dplyr::mutate(colid = paste0(cell_type, "_", donor_id)) %>%
  tibble::column_to_rownames("colid")

colnames(counts) <- rownames(coldata)
```

```{r}
coldata %>% dplyr::filter(cell_counts >= 25) %>%
  group_by(cell_type) %>%
  summarise(n = n())
```

### First process counts data via MOFA cell prep

To make the analysis consistent with other tools, we used highly variable genes from PAGODA as suggested by scITD,
however we filtered out lowly expressed genes and that could belong to a background signal.

```{r}
# Define views of interest in ct
exclude_ct <- c("Mast", "Neuronal", "prolif", "Adipo")

cts <- coldata$cell_type %>%
  unique()

cts <- cts[! cts %in% exclude_ct]

#Define HVGs coming from pagoda2 analysis
pagoda_hvgs <- readRDS("./scITDdata/hvg_list.rds")

# Filtering out background noise listed genes
# Marker genes from Kuppe et al
mrkr_genes <- read_csv("./data_MI/edgeR_cellmrkrs.csv")
mrkr_genes <- mrkr_genes %>% #dplyr::filter(!name %in% exclude_ct) %>%
  dplyr::filter(FDR < 0.01, logFC > 1) %>%
  dplyr::select(name, gene) %>%
  dplyr::rename("lineage" = name) %>%
  group_by(lineage) %>%
  nest() %>%
  dplyr::mutate(data = map(data, ~.x[[1]])) %>%
  deframe()

# Do all processing of counts matrices
pb_dat <- MOFAcellulaR::create_init_exp(counts = counts,  
                                               coldata = coldata) %>%
  MOFAcellulaR::filt_profiles(pb_dat = .,
                          cts = cts,
                          ncells = 25,
                          counts_col = "cell_counts", # This refers to the column name in testcoldata where the number of cells per profile was stored
                          ct_col = "cell_type") %>%
  MOFAcellulaR::filt_gex_byexpr(pb_dat_list = .,
                                min.count = 100,
                                min.prop = 0.25) %>%
  MOFAcellulaR::tmm_trns(pb_dat_list = .,
                         scale_factor = 1000000) %>%
  MOFAcellulaR::filt_gex_byhvg(pb_dat_list = .,
                               prior_hvg = pagoda_hvgs,
                               var.threshold = NULL) %>%
  MOFAcellulaR::filt_gex_bybckgrnd(pb_dat_list = .,
                                            prior_mrks = mrkr_genes) %>%
  MOFAcellulaR::pb_dat2MOFA(pb_dat_list = .)

saveRDS(pb_dat, "./data_MI/mi_pb_red.rds")
```

# Preparing and fitting the MOFA model 

We will try to identify 6 factors that explain the variability between patients captured by 7 different cell-types.

Every factor captures coordination of gene expression across cell types and will be called multicellular gene factors for the rest of the vignette.

It is important to clarify what these factors capture:

a) Coordinated expression of identical genes (generalistic response) across cell-types
b) Coordinated expression of different genes (cell-type specific response) across cell-types

At the same time this coordination could be associated to cell type location in the tissue, nevertheless this shouldn't be precisely true in global tissue responses.

```{r, message=FALSE}
MOFAobject <- create_mofa(pb_dat)

data_opts <- get_default_data_options(MOFAobject)

model_opts <- get_default_model_options(MOFAobject)

model_opts$spikeslab_weights <- FALSE

```

To get an optimal number of Factors, I will maximize the mean total explained variance.

So we will loop from 3 to 6 factors and observe the value

```{r, message=FALSE, eval = TRUE}
total_meanR2 <- map_chr(set_names(3:6), function(n) {
  
  print(n)
  
  model_opts$num_factors <- n
  
  train_opts <- get_default_training_options(MOFAobject)

  # Prepare MOFA model:
  MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts)
  
  outfile <- file.path("./results/MI/MOFA_mcell", 
                       paste0("MI_model", 
                       as.character(n),
                       "factors",
                       ".hdf5"))
  
  model <- run_mofa(MOFAobject, outfile)
  
  return(model@cache$variance_explained$r2_total$single_group %>% mean())

})

```


```{r, message=F, eval = TRUE}
model_opt_plt <- total_meanR2 %>%
  enframe() %>%
  dplyr::mutate(value = as.numeric(value)) %>%
  ggplot(aes(x = name, y = value)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text = element_text(size = 12)) +
  ylab("Mean total expl. var.") +
  xlab("Number of factors")

pdf("./results/MI/MOFA_mcell/model_optm.pdf", height = 2.5, width = 2.5)
plot(model_opt_plt)
dev.off()

plot(model_opt_plt)
```

# We decided to keep 6 factors for downstream analysis

```{r}
MOFA2::plot_data_overview(MOFAobject)
```

```{r}
model_outfile <- "./results/MI/MOFA_mcell/MI_model6factors.hdf5"
model <- MOFA2::load_model(model_outfile)
```

# Downstream analysis

## Association of factors to covariates

We will associate clinical covariates to factor scores. For each factor score, we will perform an analysis of variance to clinical covariates of interest. 

In this specific case we will identify which factors associate with the histological description

```{r}
meta <- sample_dict_red %>%
  dplyr::rename("sample" = "patient_region_id") %>%
  dplyr::mutate(patient_group = gsub("-enriched", "", patient_group)) %>%
  dplyr::mutate(patient_group = factor(patient_group,
                                       levels = c("myogenic", "fibrotic", "ischemic")))

write_csv(meta, "./data_MI/pb_metadata.csv")
```

### Aesthetics definitions of comparisons

```{r}
my_cond_comparisons <- list(c("myogenic", "ischemic"), c("ischemic", "fibrotic"), c("myogenic", "ischemic"))
my_batch_comparisons <- list(c("A","B"))
```

## Association to condition

ANOVA reported in the text

```{r}
expl_var_pgroup <- MOFAcellulaR::get_associations(model = model,
                                       metadata = meta,
                                        sample_id_column = "sample",
                                        test_variable = "patient_group",
                                        test_type = "categorical",
                                        categorical_type = "non-parametric",
                                        group = FALSE)

expl_var_pgroup

expl_var_pgroup %>%
  dplyr::filter(adj_pvalue <= 0.05)
```

T-test reported in the legends for factor 1

```{r}
pair_comp <- MOFAcellulaR::get_tidy_factors(model, 
                                 metadata = meta,
                                 factor = "Factor1",
                                 sample_id_column = "sample") %>%
  ggpubr::compare_means(value ~ patient_group,  
                      data = .,
                      method = "wilcox.test")

pair_comp
```

```{r, fig.height=4, fig.width=2.5}
max_val <- MOFAcellulaR::get_tidy_factors(model, 
                                 metadata = meta,
                                 factor = "Factor1",
                                 sample_id_column = "sample") %>%
  pull(value) %>%
  max() + 0.5

associated_disease <- MOFAcellulaR::get_tidy_factors(model, 
                                 metadata = meta,
                                 factor = "Factor1",
                                 sample_id_column = "sample") %>%
  pivot_wider(names_from = Factor, values_from = value) %>%
  ggplot(., aes(x = patient_group,
                y = Factor1,
                color = patient_group)) +
  geom_boxplot() +
  geom_point(size = 2) +
  ggpubr::stat_pvalue_manual(pair_comp, 
                             label = "p.adj", 
                             y.position = max_val, 
                             step.increase = 0.1,
                             tip.length = 0.01,size = 3.3) +
  theme_classic() +
  xlab("") +
  scale_color_manual(values = set_names(c("#FF6666", "#3CB371", "#4169E1"), 
                                        c("myogenic","ischemic", "fibrotic"))) +
  theme(axis.text.x = element_text(size =12, angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size =12),
        legend.position = "none") +
    theme(plot.margin = unit(c(1.0,0,0,0.5), "cm"))  #


plot(associated_disease)

pdf("./results/MI/MOFA_mcell/associated_disease.pdf", height = 2.5, width = 1.5)

plot(associated_disease)

dev.off()
```

## Association to batch

```{r}
expl_var_batch <- MOFAcellulaR::get_associations(model = model,
                                       metadata = meta,
                                        sample_id_column = "sample",
                                        test_variable = "batch",
                                        test_type = "categorical",
                                        categorical_type = "non-parametric",
                                        group = FALSE)
expl_var_batch
```

```{r}
meta$batch %>% table()
```

```{r}
my_batch_comparisons <- list(c("A","B"))

associated_batch_2 <- MOFAcellulaR::get_tidy_factors(model, 
                                 metadata = meta,
                                 factor = "Factor2",
                                 sample_id_column = "sample") %>%
  pivot_wider(names_from = Factor, values_from = value) %>%
  ggplot(., aes(x = batch,
                y = Factor2,
                color = batch)) +
  geom_boxplot() +
  geom_point(size = 2) +
  ggpubr::stat_compare_means(comparisons = my_batch_comparisons,
                             method = "wilcox.test",
                     size = 2.5) +
  theme_classic() +
  theme(axis.text.x = element_text(size =12),
        axis.text.y = element_text(size =12),
        legend.position = "none") + 
  xlab("") +
  scale_color_manual(values = set_names(c("black", "darkgrey"), c("A","B"))) 
  

plot(associated_batch_2)
```


```{r}
my_batch_comparisons <- list(c("A","B"))

associated_batch_4 <- MOFAcellulaR::get_tidy_factors(model, 
                                 metadata = meta,
                                 factor = "Factor4",
                                 sample_id_column = "sample") %>%
  pivot_wider(names_from = Factor, values_from = value) %>%
  ggplot(., aes(x = batch,
                y = Factor4,
                color = batch)) +
  geom_boxplot() +
  geom_point(size = 2) +
  ggpubr::stat_compare_means(comparisons = my_batch_comparisons,
                             method = "wilcox.test",
                     size = 2.5) +
  theme_classic() +
  theme(axis.text.x = element_text(size =12),
        axis.text.y = element_text(size =12),
        legend.position = "none") + 
  xlab("") +
  scale_color_manual(values = set_names(c("black", "darkgrey"), c("A","B"))) 
  

plot(associated_batch_4)

batch_panel <- cowplot::plot_grid(associated_batch_2, associated_batch_4, nrow = 2, align = "hv")

pdf("./results/MI/MOFA_mcell/associated_batch.pdf", height = 3, width = 1.3)

plot(batch_panel)

dev.off()
```

## Explained variance per factor and summary heatmap

We recover the explained variance per view per factor to evaluate if coordinated gene expression processes are coordinated between all the cell types or exclusive to only a group of them

```{r}
assoc_list <- list(condition = expl_var_pgroup, 
                   batch = expl_var_batch)

col_list <- list(patient_group = c("myogenic" = "#FF6666", 
                                   "ischemic" = "#3CB371", 
                                   "fibrotic" = "#4169E1"),
                 batch = c("A" = "black",
                           "B" = "darkgrey"))

scores_hmap <- plot_MOFA_hmap(model = model,
                group = FALSE,
                metadata = meta,
                sample_id_column = "sample",
                sample_anns = c("patient_group", "batch"),
                assoc_list = assoc_list,
               col_rows = col_list)

pdf("./results/MI/MOFA_mcell/MOFAcell_summ.pdf", height = 5, width = 3.7)

draw(scores_hmap)

dev.off()
```

## Dimensionality reduction

Finally, to visualize in a 2D scatter plot we use UMAP with all factors

```{r,fig.height=2.5, fig.width=4.5}
set.seed(2810)
# Factor scores per sample:
factors <- MOFA2::get_factors(model, factors = "all")[[1]] 

# UMAP of factor
factors_umap <- uwot::umap(factors, n_neighbors = 5)
colnames(factors_umap) <- c("UMAP_1", "UMAP_2")

factors_umap <- factors_umap %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(meta)

umap_plt <- ggplot(factors_umap, aes(x = UMAP_1, 
                                     y = UMAP_2, 
                                     color = patient_group,
                                     shape = batch)) +
  geom_point(size = 2.5) +
  theme_classic() +
  theme(axis.text = element_text(size =12)) +
  scale_color_manual(values = c("#FF6666", "#4169E1", "#3CB371"))

plot(umap_plt)

pdf("./results/MI/MOFA_mcell/Factor_UMAP.pdf", width = 3.8, height = 2.7)
plot(umap_plt)
dev.off()
```

We can also visualize the factors using Multidimensional scaling

```{r, fig.height=2.5, fig.width=4.5}
factors <- get_factors(model, factors = "all")[[1]] 

factors_mds <- cmdscale(dist(factors)) %>%
  as.data.frame()

colnames(factors_mds) <- c("MDS1", "MDS2")

factors_mds <- factors_mds %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(meta)

mds_plt <- ggplot(factors_mds, aes(x = MDS1, 
                                     y = MDS2, 
                                     color = patient_group,
                                     shape = batch)) +
  geom_point(size = 2.5) +
  theme_classic() +
  theme(axis.text = element_text(size =12)) +
  scale_color_manual(values = c("#FF6666", "#4169E1", "#3CB371"))

plot(mds_plt)

pdf("./results/MI/MOFA_mcell/Factor_MDS.pdf", width = 3.8, height = 2.7)
plot(mds_plt)
dev.off()
```

## Paper stats

### Recovered mean across all samples
```{r}
model@cache$variance_explained$r2_total$single_group %>% mean()
```

### Factors associated with patient group
```{r}
useful_factors <- expl_var_pgroup %>%
  dplyr::filter(adj_pvalue <= 0.05) %>%
  pull(Factor)

print(useful_factors)

model@cache$variance_explained$r2_per_factor$single_group[useful_factors,,drop = F] %>%
  colSums()

model@cache$variance_explained$r2_per_factor$single_group[useful_factors,,drop = F] %>%
  colSums() %>%
  mean()
```

### Factors associated with batch
```{r}
useful_factors <- expl_var_batch %>%
  dplyr::filter(adj_pvalue <= 0.05) %>%
  pull(Factor)

print(useful_factors)

model@cache$variance_explained$r2_per_factor$single_group[useful_factors,,drop = F] %>%
  colSums()

model@cache$variance_explained$r2_per_factor$single_group[useful_factors,,drop = F] %>%
  colSums() %>%
  mean()
```