---
title: "02.stateeval"
author: "Ricardo Ramirez"
date: "8/22/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Introduction

In here we are gonna test which genes of the observed program within a cell-type is consequence of a global response or the emergence of a cell-state

The relationship between this depends on the amount of variability of pseudobulk expressions that is associated with cell-state (from scell data) or with pure conditions

The requirements for this analysis are pseudobulk data from cell-types with annotations about condition. State pseudobulk data coming from single cell RNA seq and the loadings obtained after running MOFAcell downstream pipelines

```{r setup, include=FALSE}
library(tidyverse)
source("MOFAcell/code/MOFAcellprep.R")
```

# Recovering pseudobulk data from conditions

```{r}
cond_pb <- readRDS("./data_MI/mi_pb_red.rds") %>%
  dplyr::mutate(feature = strsplit(feature, "_") %>%
                  map_chr(., ~.x[[2]]))
```

```{r}
annotation_names <- tibble(patient_group = c("group_1", "group_2", "group_3"),
                           patient_group_name = c("myogenic-enriched", "ischemic-enriched", "fibrotic-enriched"))

```


```{r}
annotation_names <- tibble(patient_group = c("group_1", 
                                             "group_2", 
                                             "group_3"),
                           patient_group_name = c("myogenic-enriched", 
                                                  "ischemic-enriched", 
                                                  "fibrotic-enriched"))

batch_info <- read_csv("./data_MI/snrna_batch_ann.csv") %>%
  select(orig.ident, batch) %>%
  unique()

sample_dict <- read_csv("./data_MI/rna_patient_anns_revisions.csv",show_col_types = F) %>%
  left_join(annotation_names, by = "patient_group") %>%
  dplyr::select(-patient_group) %>%
  dplyr::rename("patient_group" = patient_group_name) %>%
  dplyr::mutate(patient_group = factor(patient_group, 
                                       levels = c("myogenic-enriched", "ischemic-enriched", "fibrotic-enriched"))) %>%
  #dplyr::select(-sample_id) %>%
  unique()

sample_dict_red <- sample_dict %>%
  left_join(batch_info, by = c("sample_id" = "orig.ident")) %>%
  dplyr::select(patient_region_id, major_labl, patient_group, batch) %>%
  unique() %>%
  dplyr::rename("sample" = patient_region_id)

meta <- sample_dict_red
```

Creating useful data frame

```{r}
cond_pb <- cond_pb %>%
  left_join(meta)

head(cond_pb)
```

# Recover state pseudobulk (similarly as the processing done in MOFAcell)

```{r}
state_pb <- read_csv("./data_MI/pb_data_states.csv")

colnames(state_pb)[1] <- "sample_id"

state_pb <- state_pb %>%
  column_to_rownames("sample_id") %>%
  as.matrix() %>%
  t()

state_coldata <- read_csv("./data_MI/pb_coldata_states.csv")[, -1] %>%
  column_to_rownames("colname") %>%
  dplyr::select(cell_type, cell_state, patient_region_id, patient_group, counts) %>%
  dplyr::rename("donor_id" = "patient_region_id",
                "cell_counts" = "counts")


# Define views of interest in ct
exclude_ct <- c("Mast", "Neuronal", "prolif", "Adipo", "Lymphoid", "PC", "vSMCs")

cts <- state_coldata$cell_type %>%
  unique()

cts <- cts[! cts %in% exclude_ct]

# Do all processing of counts matrices
state_pb <- create_init_exp(counts = state_pb,
                            coldata = state_coldata) %>%
  filt_profiles(pb_dat = .,
                ncells = 25,
                cts = cts) %>%
  filt_gex_byexpr(pb_dat_list = .,
                  min.count = 50,
                  min.prop = 0.1) %>%
  tmm_trns(pb_dat_list = .) %>%
  pb_dat2long(pb_dat_list = .)

head(state_pb)
```

# Import loadings

```{r}
# In this function we separate loading genes from
# positive and negative loadings, and estimate celltypes

clean_loads <- function(loadings, sep = "_") {
  
  c_load <- loadings %>%
    dplyr::mutate(ct = strsplit(celltype,"_") %>%
                    map_chr(., ~.x[1]),
                  class = strsplit(celltype,"_") %>%
                    map_chr(., ~.x[2])) %>%
    dplyr::select(gene, ct, class)
  
  return(c_load)
  
}

```


```{r}
loadings <- read_csv("./results/MI/MOFA_mcell/factor_desc/Factor1_char/loadings.csv") %>%
  clean_loads(.) %>%
  dplyr::rename("view" = "ct",
                "feature" = "gene")

head(loadings)
```

# First enrich cell-state markers into the loading matrix

```{r}
GSE_analysis = function(geneList,Annotation_DB){
  
  geneList = geneList[geneList %in% unique(unlist(Annotation_DB))]
  
  ResultsDF = matrix(0,nrow = length(Annotation_DB),ncol = 5)
  rownames(ResultsDF) = names(Annotation_DB)
  colnames(ResultsDF) = c("GenesInPathway","GenesInList","GeneNames","p_value","corr_p_value")
  
  DB_genecontent = length(unique(unlist(Annotation_DB)))
  GenesDB = DB_genecontent 
  SelectedGenes = length(geneList)
  
  for(gset in rownames(ResultsDF)){
    GP = length(Annotation_DB[[gset]])
    GL = length(intersect(Annotation_DB[[gset]],geneList))
    
    ResultsDF[gset,"GenesInList"] = GL
    ResultsDF[gset,"GenesInPathway"] = GP
    ResultsDF[gset,"GeneNames"] = paste(intersect(Annotation_DB[[gset]],geneList),collapse = ",")
    ResultsDF[gset,"p_value"] = phyper(q=GL - 1, m=GP, n=GenesDB-GP, k=SelectedGenes, lower.tail = FALSE, log.p = FALSE)
  }
  
  ResultsDF[,"corr_p_value"] = p.adjust(ResultsDF[,"p_value"],method = "BH")
  ResultsDF = data.frame(ResultsDF,stringsAsFactors = F)
  ResultsDF = ResultsDF[order(ResultsDF[,"p_value"]),]
  
  ResultsDF = ResultsDF %>% 
    rownames_to_column("gset") %>% 
    mutate_at(c("GenesInPathway","GenesInList",
                "p_value","corr_p_value"), 
              as.numeric) %>% 
    dplyr::arrange(corr_p_value,GenesInList)
  
  return(ResultsDF)
}
```

```{r}
# State markers from the Kuppe et. al. paper

state_mrkrs <- readRDS(file = "./data_MI/state_genesets_list.rds")

enrichment_per_ct <- set_names(cts) %>%
  map(., function(ct) {
    
    useful_states <- grepl(ct, names(state_mrkrs))
    useful_states <- state_mrkrs[useful_states]
    
    enrichment <- dplyr::filter(loadings, view == ct) %>%
      group_by(view,class) %>%
      nest() %>%
      dplyr::mutate(data = map(data, ~.x[[1]])) %>%
      dplyr::mutate(enrichment = map(data, GSE_analysis, Annotation_DB = useful_states)) %>%
      dplyr::select(view,class, enrichment) %>%
      unnest(cols = c(enrichment)) %>%
      dplyr::select(view, class, gset, corr_p_value)
    
    return(enrichment)
  })

plt_per_ct <- map(enrichment_per_ct, function(dat) {
  
  df_plt <- dat %>%
    dplyr::mutate(MOFA_loading = paste0(view,"_",class),
                  logpval = -log10(corr_p_value)) %>%
    dplyr::rename("state" = gset)
  
  
  enrich_plt <- ggplot(df_plt, aes(y = state, x = MOFA_loading, size = logpval)) +
    geom_point() +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 11, angle = 90, hjust = 0, vjust = 0.5),
          axis.text.y = element_text(size = 11)) +
    coord_equal()
  
  return(enrich_plt)
  
})

```

```{r, fig.height=3, fig.width= 5}

pdf("./results/MI/MOFA_mcell/factor_desc/Factor1_char/state_relationship.pdf", height = 2.5, width = 4.5)
walk(plt_per_ct, plot)
dev.off()
```

# Run state evaluation

The idea behind this evaluation is that for a given collection of pseudobulk samples of cell-states per patient, one can infer for each gene in the MOFAcell loading matrix if its gene expression is better explained by the cell-state label or the patient grouping. The biological intuition is that whilst functional cell-states emerge in case control studies upon disease contexts, it is still the case that there is a homogeneus response that is redistributed across all states.

Here we calculate for the whole matrix two types of ANOVAs, first, we calculate eta-squared scores to measure the variance explained by the patient grouping variable, and later by the state grouping variable

```{r}
estimate_statevsglobal = function(state_pb, 
                                  loadings, 
                                  ct = "CM",
                                  cond_label = "patient_group",
                                  state_label = "cell_state") {
  
  # Filter elements to ct of interest
  # and genes of interest
  
  ct_loads <- loadings %>%
    dplyr::filter(view %in% ct)
  
  ct_state <- state_pb %>%
    dplyr::filter(view %in% ct,
                  feature %in% ct_loads[["feature"]])
  
  # ct_cond <- cond_pb %>%
  #   dplyr::filter(view %in% ct,
  #                 feature %in% ct_loads[["feature"]])
  
  var1 = paste0(state_label, "_", "etasq")
  var2 = paste0(state_label, "_", "pval")
  
  # Perform ANOVA for the state label
  state_var <- ct_state %>%
    group_by(feature) %>%
    nest() %>%
    dplyr::mutate(var_expl = map(data, function(dat) {
      
      gene_aov <- aov(as.formula(paste0("value ~ ", state_label)), data = dat)
      
      aov_stats <- sjstats::anova_stats(gene_aov) %>%
        as.data.frame() %>%
        dplyr::filter(term == state_label) %>%
        dplyr::select(term, etasq, p.value)
    })) %>%
    dplyr::select(feature, var_expl) %>%
    unnest() %>%
    dplyr::arrange(desc(etasq)) %>%
    rename("etasq" = var1,
           "p.value" = var2) %>%
    dplyr::select(-term)
  
  # Perform ANOVA for the condition label
  
  var1 = paste0(cond_label, "_", "etasq")
  var2 = paste0(cond_label, "_", "pval")
  
  cond_var <- ct_state %>%
    group_by(feature) %>%
    nest() %>%
    dplyr::mutate(var_expl = map(data, function(dat) {
      
      gene_aov <- aov(as.formula(paste0("value ~ ", cond_label)), data = dat)
      
      aov_stats <- sjstats::anova_stats(gene_aov) %>%
        as.data.frame() %>%
        dplyr::filter(term == cond_label) %>%
        dplyr::select(term, etasq, p.value)
    })) %>%
    dplyr::select(feature, var_expl) %>%
    unnest() %>%
    dplyr::arrange(desc(etasq)) %>%
    rename("etasq" = var1,
           "p.value" = var2) %>%
    dplyr::select(-term)
  
  
  # Perform ANOVA for the condition label
  # cond_var <- ct_cond %>%
  #   group_by(feature) %>%
  #   nest() %>%
  #   dplyr::mutate(var_expl = map(data, function(dat) {
  #     
  #     gene_aov <- aov(as.formula(paste0("value ~ ", cond_label)), data = dat)
  #     
  #     aov_stats <- sjstats::anova_stats(gene_aov) %>%
  #       as.data.frame() %>%
  #       dplyr::filter(term == cond_label) %>%
  #       dplyr::select(term, etasq, p.value)
  #   })) %>%
  #   dplyr::select(feature, var_expl) %>%
  #   unnest() %>%
  #   dplyr::arrange(desc(etasq)) %>%
  #   rename("etasq" = var1,
  #          "p.value" = var2) %>%
  #   dplyr::select(-term)
  
  eta_sum <- left_join(state_var, cond_var, by = "feature") %>%
    ungroup() %>%
    mutate(cell_state_pval = p.adjust(cell_state_pval)) %>%
    mutate(patient_group_pval = p.adjust(patient_group_pval)) %>%
    mutate(global_ratio =patient_group_etasq/cell_state_etasq)
  
  return(eta_sum)
}

```

```{r}
plot_eta <- function(eta_sum, 
                    cond_label = "patient_group",
                    state_label = "cell_state",
                    p_thrsh = 0.01) {
  
  pval_cond <- paste0(cond_label, "_", "pval")
  pval_state <- paste0(state_label, "_", "pval")
  
  eta_sum[, pval_cond] <- eta_sum[, pval_cond] < p_thrsh
  eta_sum[, pval_state] <- eta_sum[, pval_state] < p_thrsh
  eta_sum[, "pval_both"] <- eta_sum[, pval_state] & eta_sum[, pval_cond]
  
  eta_sum[, pval_cond] <- ifelse(eta_sum[, pval_cond] == TRUE,
                                 cond_label,
                                 "")
  
  eta_sum[, pval_state] <- ifelse(eta_sum[, pval_state] == TRUE,
                                 state_label,
                                 "")
  
  eta_sum <- eta_sum %>%
    mutate(pval_both = ifelse(pval_both == TRUE,
                              "both",paste0(cell_state_pval, patient_group_pval))) %>%
     mutate(pval_both = ifelse(pval_both == "",
                              "none", pval_both)) %>%
    dplyr::rename("significance"= "pval_both")
  
  
  eta_plt <- ggplot(eta_sum, aes(x = patient_group_etasq, y = cell_state_etasq, color = significance)) +
    geom_point() +
    theme_minimal() +
    xlim(c(0,1)) +
    ylim(c(0,1)) +
    theme(axis.text = element_text(size =12),
          axis.title = element_text(size =12),
          legend.text=element_text(size=10))
  
  return(eta_plt)
  
}

```


```{r}
CM_eta <- estimate_statevsglobal(state_pb = state_pb,
                       loadings = loadings,
                       ct = "CM",
                       cond_label = "patient_group",
                       state_label = "cell_state") %>%
  mutate(ct = "CM") 

eta_plt <- CM_eta %>%
  plot_eta() +
  ylab("expl. variance by cell state") +
  xlab("expl. variance by condition") +
  guides(colour = guide_legend(override.aes = list(size=3)))


pdf("./results/MI/MOFA_mcell/factor_desc/Factor1_char/CM_eta_plt.pdf", height = 3.5, width = 4.7)

plot(eta_plt)

dev.off()

```


```{r}
Fib_eta <- estimate_statevsglobal(state_pb = state_pb,
                       loadings = loadings,
                       ct = "Fib",
                       cond_label = "patient_group",
                       state_label = "cell_state") %>%
  mutate(ct = "Fib") 

Fib_eta %>%
  plot_eta() %>%
  plot()
```

```{r}
Endo_eta <- estimate_statevsglobal(state_pb = state_pb,
                       loadings = loadings,
                       ct = "Endo",
                       cond_label = "patient_group",
                       state_label = "cell_state") %>%
  mutate(ct = "Endo") 

Endo_eta %>%
  plot_eta() %>%
  plot()
```

```{r}
Myeloid_eta <- estimate_statevsglobal(state_pb = state_pb,
                       loadings = loadings,
                       ct = "Myeloid",
                       cond_label = "patient_group",
                       state_label = "cell_state") %>%
  mutate(ct = "Myeloid") 

Myeloid_eta %>%
  plot_eta() %>%
  plot()
```

# Global responses across cell types

```{r}
global_ratio_plt <- bind_rows(CM_eta, Fib_eta, 
          Endo_eta, Myeloid_eta) %>%
  ggplot(aes(x = ct, y = log2(global_ratio))) +
  geom_violin() +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size =12)) +
  xlab("") +
  ylab("log2(condition/state)")

plot(global_ratio_plt)

pdf("./results/MI/MOFA_mcell/factor_desc/Factor1_char/eta_ratio_plt.pdf", height = 3.5, width = 3.5)

plot(global_ratio_plt)

dev.off()

```

# Examples of genes 

# NFATC4

```{r}
CM_eta %>%
  dplyr::filter(feature == "NFATC4")
```


```{r}
state_ex_plt <- state_pb %>%
  dplyr::filter(view == "CM",
                feature == "NFATC4") %>%
  ggplot(aes(x = cell_state, y = value)) +
  geom_boxplot() +
  geom_point() +
  theme_minimal() +
  ylab("expression") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size =12))

cond_ex_plt <- state_pb %>%
  dplyr::filter(view == "CM",
                feature == "NFATC4") %>%
  left_join(annotation_names) %>%
  ggplot(aes(x = patient_group_name, y = value)) +
  geom_boxplot() +
  geom_point() +
  theme_minimal() +
  ylab("expression") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size =12))

expression_plt <- cowplot::plot_grid(cond_ex_plt, state_ex_plt, align = "hv", ncol = 2,rel_widths = c(0.75,1))

pdf("./results/MI/MOFA_mcell/factor_desc/Factor1_char/CM_NFATC4_example.pdf", height = 3.5, width = 3.5)

plot(expression_plt)

dev.off()

```

# MYH7

```{r}
CM_eta %>%
  dplyr::filter(feature == "MYH7")
```

```{r}
state_pb %>%
  dplyr::filter(view == "CM",
                feature == "MYH7") %>%
  ggplot(aes(x = cell_state, y = value)) +
  geom_boxplot() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 12))

state_pb %>%
  dplyr::filter(view == "CM",
                feature == "MYH7") %>%
  left_join(annotation_names) %>%
  ggplot(aes(x = patient_group_name, y = value)) +
  geom_boxplot() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 12))
```




